---
title: "Investigating the 191 PDZ factors"
author: "Joonsuk Kang"
date: "3/31/2020"
output: html_document
---

This article investigate the 191 factors for PDZ domain data fitted with point normal flashier in [Tackling Low-rankness of Estimated Precision Matrix](lowrank.html) 

# Data

```{r}
library(tidyverse); library(flashier); library(RColorBrewer); library(gridExtra)
data.raw <- read_table2("data/PDZ", col_names = FALSE)
data.raw %>% mutate_if(is.numeric,as.factor) -> data
fastDummies::dummy_cols(data, remove_selected_columns=TRUE) -> data
data <- as.matrix(data)

fit <- readRDS("output/fit_flashier_f191bf.rds")
f <- t(fit$loadings.pm[[2]])
col.idx <- matrix(as.numeric(unlist(strsplit(substr(colnames(f),2,100), "_"))), byrow=TRUE, ncol=2)
```

# Sample Figure

Each heatmap represents each of the 191 factors: position as rows (1--89) and amino acid (states) as columns (1--21).

As a sample, the first factor is shown below. For visualization, values greater than 0.22 were fixed to 0.22 and likewise for those smaller than -0.22. Note that a matrix corresponding to a factor has Frobenius norm 1. Grey cells are (position, amino acid) pair which has zero appearance in the whole data. 

The first factor sort of captures the mean trend, and almost every term is positive (the smallest one is -0.0010). Along with the first factor, the proportion of amino acid per position represented in the data is shown. As we can check in the following scatterplot, the weight in the first factor is closely related to the mean trend.


```{r}
breaksList = seq(0,1,by=0.05)
colpal <- colorRampPalette(brewer.pal(n =9, name = "Oranges"))(length(breaksList))
colpal[1] <- "#ffffff"

colSums(data) -> data.count
mat.dc <- matrix(NA, nrow=89, ncol=21)
for (j in 1:ncol(f)){   mat.dc[col.idx[j,1], col.idx[j,2]] <- data.count[j]/nrow(data)  }

plot_list1=list()
pheatmap::pheatmap(mat.dc, cluster_rows=FALSE, cluster_cols=FALSE, cellwidth=5,cellheight=5,
                   main="Proprtion of amino acid per position",
                   color=colpal, border_color="grey95",
                   breaks = breaksList, silent=TRUE
                   ) -> temp.fig
plot_list1[[1]] <- temp.fig[[4]]
```


```{r factor1, fig.height=7}
breaksList = seq(-0.22, 0.22, by = 0.04)
colpal <- colorRampPalette(rev(brewer.pal(n =10, name = "RdBu")))(length(breaksList))
colpal[6] <- "#ffffff"

i <- 1
one.f <- f[i,]; one.f[one.f>0.22] <- 0.22; one.f[one.f<-0.22]<- -0.22
mat.f <- matrix(NA,nrow=89,ncol=21)
for (j in 1:ncol(f)){   mat.f[col.idx[j,1], col.idx[j,2]] <- one.f[j]  }
pheatmap::pheatmap(mat.f, cluster_rows=FALSE, cluster_cols=FALSE, cellwidth=5,cellheight=5,
                   main=paste0("Factor ", i),
                   color=colpal, border_color="grey95",
                   breaks = breaksList, silent=TRUE
                   ) -> temp.fig
plot_list1[[2]] <- temp.fig[[4]]
grid.arrange(arrangeGrob(grobs= plot_list1,ncol=2))
```


```{r comparison}
data.frame(data = c(mat.dc),
           factor1 = c(mat.f)) %>%
  ggplot()+geom_point(aes(x=data, y=factor1), alpha=0.1)+
  ggtitle("Proportion of Amino Acid per Position vs Weight in Factor 1")
```

# Count

We counted the number of strong signals in each factor using two cutoffs: absolute size 0.05 and relative size (largest 100 values). A factor typically has zero or one strong signal associated with a position, sometimes two signals, and rarely more than that.


### Cutoff: size 0.05

We classified values larger than 0.05 as 1 and 0 otherwise. Then, for each factor, count the number of 1's in each position. Each column corresponds to one of the 191 factors and each row to one of the 89 positions.

Note: 0=white, 1=blue, 2=red, 3=green, 4=yellow, 5=black (maximum)

```{r cutoff1, fig.height=7, fig.width=15}
# construct binary version of factors with cutoff 0.05
f.bin <- f
f.bin[abs(f.bin)<0.05] <- 0
f.bin[f.bin!=0] <- 1

# create count data
f.count <- matrix(NA, nrow=89, ncol=nrow(f))
for (i in 1:nrow(f)){
  
    one.f <- f.bin[i,]
    mat.f <- matrix(NA,nrow=89,ncol=21)
    for (j in 1:ncol(f)){   mat.f[col.idx[j,1], col.idx[j,2]] <- one.f[j]  }
    f.count[,i] <- rowSums(mat.f, na.rm=TRUE) # number of amino acids (states) with cell value greater than 0.05
}

# plot
colpal2 <- c('white', 'blue', 'red', 'green', 'yellow', 'black')
pheatmap::pheatmap(f.count, cluster_rows=FALSE, cluster_cols=FALSE,
                   cellwidth=5, cellheight=5,
                   color=colpal2, border_color="grey95"
                   )
```

### Cutoff: top 100

Alternatively, use relative threshold (top 100) instead of absolute threshold of 0.05. 

For each factor, classify top 100 largest values as 1 and 0 otherwise. Then, for each factor, count the number of 1's in each position. Each column corresponds to one of the 191 factors and each row to one of the 89 positions.

Note: 0=white, 1=blue, 2=red, 3=green, 4=yellow, 5=black, 6=orange, 7=pink (maximum)



```{r cutoff2, fig.height=7, fig.width=15}
f.t100 <- abs(f)

# create count data
f.count <- matrix(NA, nrow=89, ncol=nrow(f))
for (i in 1:nrow(f)){
  
    one.f <- f.t100[i,]
    top100 <- as.numeric(sort(one.f, decreasing=TRUE)[100])
    if(top100>0){one.f <- ifelse(one.f>=top100, 1, 0)}
    if(top100==0){one.f <- ifelse(one.f>0, 1, 0)}
    
    mat.f <- matrix(NA,nrow=89,ncol=21)
    for (j in 1:ncol(f)){   mat.f[col.idx[j,1], col.idx[j,2]] <- one.f[j]  }
    f.count[,i] <- rowSums(mat.f, na.rm=TRUE) # number of amino acids (states) with cell value greater than 0.05
}

# plot
#breaksList2 = seq(0,8,by=1)
colpal2 <- c('white', 'blue', 'red', 'green', 'yellow', 'black', 'orange', 'pink')

pheatmap::pheatmap(f.count, cluster_rows=FALSE, cluster_cols=FALSE,
                   cellwidth=5, cellheight=5,
                   color=colpal2, border_color="grey95"
                   )
```



# The Catalog

Here, the heatmap for all the 191 factors is catalogued.

```{r catalog, fig.width=10, fig.height=80}
plot_list2=list()
for (i in 1:nrow(f)){
  one.f <- f[i,]; one.f[one.f>0.22] <- 0.22; one.f[one.f<-0.22]<- -0.22
  mat.f <- matrix(NA,nrow=89,ncol=21)
  for (j in 1:ncol(f)){   mat.f[col.idx[j,1], col.idx[j,2]] <- one.f[j]  }
  temp.fig <- pheatmap::pheatmap(mat.f, cluster_rows=FALSE, cluster_cols=FALSE, cellwidth=3,cellheight=3,
                     main=paste0("Factor ", i),
                     color=colpal, border_color="grey95",
                     breaks = breaksList, legend=FALSE, 
                     silent=TRUE
                     )
  plot_list2[[i]] = temp.fig[[4]] 
}
grid.arrange(arrangeGrob(grobs= plot_list2,ncol=10))
```

