---
title: "flashier on toy data: precision matrix estimation"
author: "Joonsuk Kang"
date: "2020-03-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---



## Data

Matrix $A$ with $dim(A)=N \times L$ and $A_{ij}\in \{1,2,\dots,q\}$ 
where N=300 sequences; L=20 positions; q=10 potts-states

```{r}
library(tidyverse); library(tictoc); library(pheatmap)
data.raw <- read_table2("data/align300_q10", col_names = FALSE)
head(data.raw)

# change data type: from numeric to factor
data.raw %>% mutate_if(is.numeric,as.factor) -> data

# change data type: from factor to binary dummies
fastDummies::dummy_cols(data, remove_selected_columns=TRUE) -> data
data <- as.matrix(data)

# save column index as 2-column (position, state) matrix
col.idx <- matrix(as.numeric(unlist(strsplit(substr(colnames(data),2,100), "_"))), byrow=TRUE, ncol=2)
```


## Functions

#### `get_CPM`: function to obtain compressed precision matrix (CPM)

```{r}
# input: L, F, E     where matrix X=LF+E
#        data column index as [position, state] matrix
# output: compressed precision matrix
get_CPM <- function(l,f,e, col.idx){
  
    # exclude first factor (which captures mean level) and loadings for numerical stability
    #apply(l,2,sd) # check standard deviation of the loadings by factors
    l2 <- l[,-1]
    f2 <- f[-1,]
    
    Psi <- cov(e) # error covariance
    Psi.inv <- diag(diag(Psi)^{-1})
    
    Lambda.L <- cov(l2)
    Lambda.L.inv <- diag(diag(Lambda.L)^{-1})
    Omega <- Psi.inv - Psi.inv %*% t(f2) %*% solve(Lambda.L.inv+f2%*%Psi.inv%*%t(f2)) %*% f2 %*% Psi.inv
  
    # measure position i -- position j interaction as sqrt(sum of squares of Omega_{k, l})
    #         where position(k)=i and position(l)=j
    
    data.frame(value = c(Omega),
               position1 = rep(col.idx[,1], times=nrow(col.idx)),
               position2 = rep(col.idx[,1], each=nrow(col.idx))
               ) %>%
      group_by(position1, position2) %>%
      summarise(value = sqrt(sum(value^2))) -> sumsq
    
    # compressed precision matrix
    matrix(sumsq$value, byrow=FALSE, 
           ncol=length(unique(col.idx[,1])) # = number of positions
           ) -> CPM
    
    return(CPM)
}
```


#### `plot_CPM`: function to plot compressed precision matrix (CPM)

```{r}
# input: compressed precision matrix (output from `get_CPM`), 
#        data column index as [position, state] matrix
#        cutoff: to make the figure to easy to check, do not show cells with values < cutoff
# output: heatmap of compressed precision matrix (dim = #positions X #positions)

plot_CPM <- function(CPM, col.idx, cutoff=0){
  
    # plot only off-diagonal elements (otherwise, diagonal elements dominate visually)
    diag(CPM) <- NA
    CPM[CPM<cutoff] <- NA
    
    pheatmap(CPM, cluster_rows=FALSE, cluster_cols=FALSE,
             main="Compressed Precision Matrix",
             display_numbers=TRUE)
}
```


## Estimations

```{r}
# maximum number of factors to be added
fmax <- 50
```


### flash: point normal prior; column specific residual variance

```{r flash}
# fit models
tic()
fit <- flashr::flash(data, backfit=TRUE, greedy=TRUE, verbose=FALSE, Kmax=fmax)
toc()

# calculate l, f, e
l <- flashr::flash_get_ldf(fit)$l
f <- t(flashr::flash_get_ldf(fit)$f)
e <- data - (l %*% f) 

# obtain compressed precision matrix
CPM <- get_CPM(l, f, e, col.idx)

# plot
plot_CPM(CPM)
```


### flashier: point normal prior; column specific residual variance

```{r flashier_pn}
library(flashier)
tic()
fit <- flashier::flash(data, backfit=TRUE, verbose.lvl = 0, greedy.Kmax = fmax, 
                       prior.family = prior.point.normal(),
                       var.type=2
                       ) 
toc()

# calculate l, f, e
l <- fit$loadings.pm[[1]]
f <- t(fit$loadings.pm[[2]])
e <- data - (l %*% f) 

# obtain compressed precision matrix
CPM <- get_CPM(l, f, e, col.idx)

# plot
plot_CPM(CPM)
```


### flashier: nonnegative prior; column specific residual variance

```{r flashier_nn}
library(flashier)
tic()
fit <- flashier::flash(data, backfit=TRUE, verbose.lvl = 0, greedy.Kmax = fmax, 
                       prior.family = prior.nonnegative(),
                       var.type=2
                       ) 
toc()

# calculate l, f, e
l <- fit$loadings.pm[[1]]
f <- t(fit$loadings.pm[[2]])
e <- data - (l %*% f) 

# obtain compressed precision matrix
CPM <- get_CPM(l, f, e, col.idx)

# plot
plot_CPM(CPM)
```


## Summary

As expected, the two compressed precision matrices from `flashr::flash` and `flashier::flash` are almost identical. Flashier result with nonnegative prior has greater noise (especially the (1--10)X(15--20) blocks) while the size of signals have decreased.
