---
title: "flash on toy data"
author: "Joonsuk Kang"
date: "2020-03-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Data

Matrix $A$ with $dim(A)=N \times L$ and $A_{ij}\in \{1,2,\dots,q\}$ 
where N=300 sequences; L=20 positions; q=10 potts-states

```{r}
library(tidyverse)
data <- read_table2("data/align300_q10", col_names = FALSE)
```

Data in original format:
```{r}
head(data)
```

Data in binary coding:
```{r}
# change data type: from numeric to factor
data %>% mutate_if(is.numeric,as.factor) -> data

library(fastDummies)
fastDummies::dummy_cols(data, remove_selected_columns=TRUE) -> data
data <- as.matrix(data)
```



## flash

```{r}
library(flashr)
f.gb = flash(data, backfit=TRUE, greedy=TRUE, verbose=FALSE)
ldf.gb = flash_get_ldf(f.gb)
```


The fitted model has 44 factors; the values of the factors are shown below:

```{r}
dim(ldf.gb$f) # 200 X 44: 44 factors with length 200 (20 X 10) each
data.frame(factor = rep(1:44, each=200),
           position = rep(rep(1:20, each=10), times=44),
           state = rep(1:10, times=20*44),
           value = c(ldf.gb$f)) -> f
```


Columns correspond to factors and rows to positions. The figure in cell (i,j) illustrates the factor j's value for 10 states (from 1 on the far left to 10 on the far right) on position i.

The first factor captures the mean value. Also, we can find two clear patterns

+ ten factors (2--10, 12) put an almost identical weight on one state for positions 15--20; and 0 for all the other sattes and positions
+ ten factors (11, 13--21) put an almost identical weight on one state for positions 11-14; and 0 for all the other sattes and positions

```{r, fig.width=20, fig.height=20}
f %>%
  ggplot()+geom_line(aes(x=state, y=value))+facet_grid(position~factor)
```



### Factors 2--10 and 12

A closer look at the ten factors (2--10, 12) and positions (15--20)
```{r}
f %>%
  filter((factor>=2 & factor<=10)|factor==12) %>%
  filter(position>=15) %>%
  ggplot(aes(x=state, y=value))+geom_line(alpha=0.2)+
  geom_point()+
  facet_grid(position~factor, labeller=labeller(position=label_both, factor=label_both))

f %>%
  filter((factor>=2 & factor<=10)|factor==12) %>%
  filter(position>=15) %>%
  group_by(factor, position) %>%
  arrange(-abs(value)) %>%
  filter(row_number()==1) %>%
  ungroup() %>% arrange(factor, position) %>% select(-value) -> f.15to20

# the state with highest weight for factors/positions 
xtabs(state ~ position+factor, data=f.15to20)
```

Each of the ten factors assigns a strong weight to each state (of 10 states) for positions 15--20 and 0 otherwise.


### Factors 11 and 13--21

A closer look at the ten factors (11, 13--21) and positions (11--14)
```{r}
f %>%
  filter((factor>=13 & factor<=21)|factor==11) %>%
  filter(position>=11 & position<=14) %>%
  ggplot(aes(x=state, y=value))+geom_line(alpha=0.2)+
  geom_point()+
  facet_grid(position~factor, labeller=labeller(position=label_both, factor=label_both))

f %>%
  filter((factor>=13 & factor<=21)|factor==11) %>%
  filter(position>=11 & position<=14) %>%
  group_by(factor, position) %>%
  arrange(-abs(value)) %>%
  filter(row_number()==1) %>%
  ungroup() %>% arrange(factor, position) %>% select(-value) -> f.11to14

# the state with highest weight for factors/positions 
xtabs(state ~ position+factor, data=f.11to14)
```
Each of the ten factors assigns a strong weight to each state (of 10 states) for positions 11--14 and 0 otherwise.


### More general pattern: what positions are related

While previous two results show a clear and deterministic pattern of states among positions, a general estimation of interest would be the correlation between positions. 

One way to estimate whether two positions are correlated is counting the number of factors which has large weights on both pair of positions.

```{r}
f %>%
  filter(factor!=1) %>% # exclude first factor capturing mean level
  group_by(factor, position) %>% 
  summarise(sumsq = sum(value^2)) %>%
  ggplot()+
  geom_tile(aes(x=factor, y=position, fill=sumsq), col='white')+
  scale_y_reverse()+
  ggtitle("Sum of squares of weights within a position")

f %>%
  filter(factor!=1) %>% # exclude first factor capturing mean level
  group_by(factor, position) %>% 
  summarise(sumsq = sum(value^2)) %>%
  mutate(bigweight=factor(ifelse(sumsq>0.1, 1, 0))) -> f.bw

f.bw %>%
  ggplot()+
  geom_tile(aes(x=factor, y=position, fill=bigweight), 
            col='white', alpha=0.5)+
  scale_y_reverse()+
  ggtitle("1 if sum of squares of weights within a position > 0.1")

```

```{r}
f.bw %>% filter(bigweight==1) %>% select(factor, position) -> f.bw

find_strength <- function(i){
    f.bw %>% 
      filter(position==i) %>%
      rename(position1=position) %>%
      inner_join(f.bw, by='factor') %>%
      rename(position2=position) %>%
      filter(position1!=position2) %>%
      group_by(position1, position2) %>% 
      summarise(n=n())
}

positions_strength <- find_strength(1)
for (i in 2:20){
  positions_strength <- rbind(positions_strength,
                              find_strength(i))
}
positions_strength %>%
  ggplot()+
  geom_text(aes(x=position1, y=position2, label=n))+
  scale_y_reverse()+scale_x_reverse()+
  ggtitle("number of factors with large weights on position i and j")

positions_strength %>% 
  filter(n>1) %>%
  ggplot()+
  geom_text(aes(x=position1, y=position2, label=n))+
  scale_y_reverse()+scale_x_reverse()+
    ggtitle("number of factors with large weights on position i and j: excluding 1s")
```



## Summary


`flash` successfully recovers the structure from the toy data 

1) **what** positions are related: (15,16,17,18,19,20), (11,12,13,14), (2,5), (4,7), (6,9)
2) **how** the positions are related: there was a clear deterministic pattern for the positions 15--20 and 11--14, but needs more investigation for the other related positions



