---
title: "flashier on toy data"
author: "Joonsuk Kang"
date: "2020-04-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Data

Matrix $A$ with $dim(A)=N \times L$ and $A_{ij}\in \{1,2,\dots,q\}$ 
where N=300 sequences; L=20 positions; q=10 potts-states

```{r}
library(tidyverse)
data <- read_table2("data/align300_q10", col_names = FALSE)

# change data type: from numeric to factor
data %>% mutate_if(is.numeric,as.factor) -> data

# change data type: from factor to binary dummies
library(fastDummies)
fastDummies::dummy_cols(data, remove_selected_columns=TRUE) -> data
data <- as.matrix(data)
```



## Baseline: results from `flash`

Previously, the estimated model had 44 factors. Here, we set the maximum number of factors as 30 to make comparison with other estimation results easier.

```{r, fig.width=20, fig.height=20}
#library(flashr)
library(tictoc)
tic()
fit <- flashr::flash(data, backfit=TRUE, greedy=TRUE, verbose=FALSE,
                     Kmax=30) # at most 30 factors
toc()
factor <- flashr::flash_get_ldf(fit)$f
l <- ncol(factor)
data.frame(factor = rep(1:l, each=200),
           position = rep(rep(1:20, each=10), times=l),
           state = rep(1:10, times=20*l),
           value = c(factor)) %>%
  ggplot()+geom_line(aes(x=state, y=value))+facet_grid(position~factor)+
  ggtitle("Model: flash")

# adjust the sign of factors so that the mean weight for each factor is positive
for (colidx in 1:ncol(factor)){
  s <- sign(colMeans(factor))[colidx]
  factor[,colidx] <- factor[,colidx]*s
}

data.frame(factor = rep(1:l, each=200),
           position = rep(rep(1:20, each=10), times=l),
           state = rep(1:10, times=20*l),
           value = c(factor)) %>%
  ggplot()+geom_line(aes(x=state, y=value))+facet_grid(position~factor)+
  ggtitle("Model: flash, sign adjusted")

```

## Estimation with `flashier` package

The `flashier` package provides a variety of choices for prior distribution families. The results from point normal and nonnegative priors are shown below. Note that the `flash` package has point normal prior family as default option. Also, column specific residual variances are assumed in both flash and flashier.

#### point normal prior (default)

The default for prior family is point normal: `prior.family = prior.point.normal()`

```{r, fig.width=20, fig.height=20}
library(flashier)
tic()
fit <- flashier::flash(data, greedy.Kmax = 30, backfit=TRUE, # at most 30 factors
                       verbose.lvl = 0,
                       var.type=2 # column specific residual variance
                       ) 
toc()
factor <- fit$loadings.pm[[2]]
l <- ncol(factor)

# adjust the sign of factors so that the mean weight for each factor is positive
for (colidx in 1:ncol(factor)){
  s <- sign(colMeans(factor))[colidx]
  factor[,colidx] <- factor[,colidx]*s
}

data.frame(factor = rep(1:l, each=200),
           position = rep(rep(1:20, each=10), times=l),
           state = rep(1:10, times=20*l),
           value = c(factor)) %>%
  ggplot()+geom_line(aes(x=state, y=value))+facet_grid(position~factor)+
  ggtitle("Model: flashier with point normal prior, sign adjusted")

```

For the same setting (point normal prior and column specific residual variance), the two approaches show almost idencial resuts. The flashier implementation is two times faster. 


#### nonnegative prior 

```{r, fig.width=20, fig.height=20}
#library(flashier)
tic()
fit <- flashier::flash(data, greedy.Kmax = 30, backfit=TRUE,
                       prior.family = prior.nonnegative(),
                       verbose.lvl=0,
                       var.type=2 # column specific residual variance
                       ) 
toc()
factor <- fit$loadings.pm[[2]]
l <- ncol(factor)
data.frame(factor = rep(1:l, each=200),
           position = rep(rep(1:20, each=10), times=l),
           state = rep(1:10, times=20*l),
           value = c(factor)) %>%
  ggplot()+geom_line(aes(x=state, y=value))+facet_grid(position~factor)+
  ggtitle("Model: flashier with nonnegative prior")
```

The fitted model with nonnegative prior has smaller number of factors (18) than the other fitted models. The model captures the two strong correlation structures in positions 11-14 and 15--20, and also correlations between (2 and 5) and (6 and 9).


#### nonnegative L (loading) and point normal F (factor)


```{r nnpn, fig.width=20, fig.height=20}
#library(flashier)
tic()
fit <- flashier::flash(data, greedy.Kmax = 30, backfit=TRUE,
                       prior.family = c(prior.nonnegative(), prior.point.normal()),
                       verbose.lvl=0,
                       var.type=2 # column specific residual variance
                       ) 
toc()
factor <- fit$loadings.pm[[2]]
l <- ncol(factor)
data.frame(factor = rep(1:l, each=200),
           position = rep(rep(1:20, each=10), times=l),
           state = rep(1:10, times=20*l),
           value = c(factor)) %>%
  ggplot()+geom_line(aes(x=state, y=value))+facet_grid(position~factor)+
  ggtitle("Model: flashier with nonnegative L and point normal F prior")
```
