---
title: "Precision Matrix Estimation: PDZ domain"
author: "Joonsuk Kang"
date: "2020-03-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

**Task**: Precision Matrix Estimation with PDZ domain data.

For the estimation details, refer to ["Precision Matrix Estimation using flash"](https://joonsukkang.github.io/aaint/flash_omega.html).



## Data

Matrix $A$ with $dim(A)=N \times L$ and $A_{ij}\in \{1,2,\dots,q\}$ 
where N=1692 sequences; L=89 positions; q=21 potts-states

```{r}
library(tidyverse); library(tictoc); library(pheatmap)
data.raw <- read_table2("data/PDZ", col_names = FALSE)
head(data.raw)

# change data type: from numeric to factor
data.raw %>% mutate_if(is.numeric,as.factor) -> data

# change data type: from factor to binary dummies
fastDummies::dummy_cols(data, remove_selected_columns=TRUE) -> data
data <- as.matrix(data)

# save column index as 2-column (position, state) matrix
col.idx <- matrix(as.numeric(unlist(strsplit(substr(colnames(data),2,100), "_"))), byrow=TRUE, ncol=2)
```



## Functions

#### `get_CPM`: function to obtain compressed precision matrix (CPM)

```{r}
# input: L, F, E     where matrix X=LF+E
#        data column index as [position, state] matrix
# output: compressed precision matrix
get_CPM <- function(l,f,e, col.idx){
  
    # exclude first factor (which captures mean level) and loadings for numerical stability
    #apply(l,2,sd) # check standard deviation of the loadings by factors
    l2 <- l[,-1]
    f2 <- f[-1,]
    
    Psi <- cov(e) # error covariance
    Psi.inv <- diag(diag(Psi)^{-1})
    
    Lambda.L <- cov(l2)
    Lambda.L.inv <- diag(diag(Lambda.L)^{-1})
    Omega <- Psi.inv - Psi.inv %*% t(f2) %*% solve(Lambda.L.inv+f2%*%Psi.inv%*%t(f2)) %*% f2 %*% Psi.inv
  
    # measure position i -- position j interaction as sqrt(sum of squares of Omega_{k, l})
    #         where position(k)=i and position(l)=j
    
    data.frame(value = c(Omega),
               position1 = rep(col.idx[,1], times=nrow(col.idx)),
               position2 = rep(col.idx[,1], each=nrow(col.idx))
               ) %>%
      group_by(position1, position2) %>%
      summarise(value = sqrt(sum(value^2))) -> sumsq
    
    # compressed precision matrix
    matrix(sumsq$value, byrow=FALSE, 
           ncol=length(unique(col.idx[,1])) # = number of positions
           ) -> CPM
    
    return(CPM)
}
```


#### `plot_CPM`: function to plot compressed precision matrix (CPM)

```{r}
# input: compressed precision matrix (output from `get_CPM`), 
#        cutoff: to make the figure to easy to check, do not show cells with values < cutoff
# output: heatmap of compressed precision matrix (dim = #positions X #positions)

plot_CPM <- function(CPM, cutoff=0, type=1){
  
    # plot only off-diagonal elements (otherwise, diagonal elements dominate visually)
    diag(CPM) <- NA
    CPM[CPM<cutoff] <- NA
    
    if(type==1){pheatmap(CPM, cluster_rows=FALSE, cluster_cols=FALSE,
             main="Compressed Precision Matrix")}
    if(type==2){pheatmap(sqrt(CPM), cluster_rows=FALSE, cluster_cols=FALSE,
              main="Compressed Precision Matrix: sqrt transformed")}
    if(type==3){pheatmap(log(CPM), cluster_rows=FALSE, cluster_cols=FALSE,
                    main="Compressed Precisoin Matrix: log transformed")}
}

```





## Estimations

```{r}
# maximum number of factors to be added
fmax <- 50
```


### flash (point normal)
```{r, eval=FALSE}
# fit models
tic()
fit <- flashr::flash(data, backfit=TRUE, greedy=TRUE, verbose=FALSE, Kmax=fmax)
toc() # 23709.384 sec elapsed
saveRDS(fit, "output/fit_flash.rds")
```


```{r flash, fig.width=7, fig.height=7}
fit <- readRDS("output/fit_flash.rds")

# calculate l, f, e
l <- flashr::flash_get_ldf(fit)$l
f <- t(flashr::flash_get_ldf(fit)$f)
e <- data - (l %*% f) 

# obtain compressed precision matrix
CPM <- get_CPM(l, f, e, col.idx)

# plot
plot_CPM(CPM)
```


### flashier: point normal prior
```{r, eval=FALSE}
tic()
fit <- flashier::flash(data, backfit=TRUE, verbose.lvl = 0, greedy.Kmax = fmax, 
                       prior.family = prior.point.normal(),
                       var.type=2) # column-specific error variance
toc() # 282.459 sec elapsed
saveRDS(fit, "output/fit_flashier.rds")
```


```{r flashier_pn, fig.width=7, fig.height=7}
fit <- readRDS("output/fit_flashier.rds")

# calculate l, f, e
l <- fit$loadings.pm[[1]]
f <- t(fit$loadings.pm[[2]])
e <- data - (l %*% f) 

# obtain compressed precision matrix
CPM <- get_CPM(l, f, e, col.idx)

# plot
plot_CPM(CPM)
```


#### flashier: nonnegative prior
```{r, eval=FALSE}
library(flashier)
tic()
fit <- flashier::flash(data, backfit=TRUE, verbose.lvl = 0, greedy.Kmax = fmax, 
                       prior.family = prior.nonnegative(),
                       var.type=2) # column-specific error variance
toc() # 5400.038 sec elapsed
saveRDS(fit, "output/fit_flashier_nn.rds")
```


```{r flashier_nn, fig.width=7, fig.height=7}
fit <- readRDS("output/fit_flashier_nn.rds")

# calculate l, f, e
l <- fit$loadings.pm[[1]]
f <- t(fit$loadings.pm[[2]])
e <- data - (l %*% f) 

# obtain compressed precision matrix
CPM <- get_CPM(l, f, e, col.idx)

# plot
plot_CPM(CPM)
```

## Summary

With the point normal prior, `flashier` is much faster than the `flash` implementation while resulting in quite similar figure. A slight difference is that `flashier` has fewer strong signals: the four yellow dots and their neighbors around the middle shown in flash result are all gone. But big structures with strong signals are robust. The figure with nonnegative prior shows less patterns. 

We cannot tell from this data that the captured patterns (and the differences among these three patterns) are signal or noise. Side information on the protein structure will be able to measure the performance of these.

For visual inspection, the figure for point normal flashier result is shown in 1) original scale, 2) square-root scale, and 3) log scale.

```{r 3figs, fig.width=7, fig.height=7}
fit <- readRDS("output/fit_flashier.rds")

# calculate l, f, e
l <- fit$loadings.pm[[1]]
f <- t(fit$loadings.pm[[2]])
e <- data - (l %*% f) 

# obtain compressed precision matrix
CPM <- get_CPM(l, f, e, col.idx)

# plot
plot_CPM(CPM)
plot_CPM(CPM, type=2)
plot_CPM(CPM, type=3)
```
