---
title: "Fitting the Toy Data of Ingraham and Marks (2017, ICML)"
author: "Joonsuk Kang"
date: "2020-05-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

**Task**: Precision Matrix Estimation with toy data from Ingraham and Marks (2017, ICML) ["Variational inference for sparse and undirected models"](https://arxiv.org/abs/1602.03807)

For the estimation details, refer to ["Precision Matrix Estimation using flash"](https://joonsukkang.github.io/aaint/flash_omega.html).

The paper's code could be found here: https://github.com/debbiemarkslab/persistent-vi


## Data

Matrix $A$ with $dim(A)=N \times L$ and $A_{ij}\in \{1,2,\dots,q\}$ 
where N=400 sequences; L=50 positions; q=20 potts-states

```{r}
library(tidyverse); library(tictoc); library(pheatmap); library(gridExtra); library(RColorBrewer)
library(R.matlab)

data.raw <- readMat("https://github.com/debbiemarkslab/persistent-vi/blob/master/potts/original_dataset.mat?raw=true")
truth.location <- data.raw$C
data.raw <- data.raw$sample[1:400,] # following the paper, used first 400 observations (sequences) out of the 2000 sample

# change data type: from numeric to factor
data.frame(data.raw) %>% mutate_if(is.integer,as.factor) -> data

# change data type: from factor to binary dummies
fastDummies::dummy_cols(data, remove_selected_columns=TRUE) -> data
data <- as.matrix(data)

# save column index as 2-column (position, state) matrix
col.idx <- matrix(as.numeric(unlist(strsplit(substr(colnames(data),2,100), "_"))), byrow=TRUE, ncol=2)
```

### Data Visualization

To begin with, let's visualize the data. The figure below shows the frequency of states at each of the first ten positions.

```{r}
data.frame(position = rep(1:10,each=400),
           state = as.factor(data.raw[,1:10])) %>% group_by(position, state) %>%
  summarise(count=n()) %>% 
  ggplot()+
  geom_col(aes(x=state, y=count))+
  geom_abline(slope=0, intercept=20, col='blue')+
  ggtitle("Frequency of States at the first position")+
  facet_wrap(~position)
```

These heatmaps show the frequency for the pair of states at positions 1 and 2, or positions 3 and 4.

```{r, fig.width=7, fig.height=7}
idx_x <- 1 # corresponding to the position going into the x axis
idx_y <- 2 # corresponding to the position going into the y axis

mat.temp <- matrix(0, nrow=20, ncol=20)
for (i in 1:400){
  val_x <- data.raw[i,idx_x] 
  val_y <- data.raw[i,idx_y] 
  mat.temp[val_y,val_x] <- mat.temp[val_y,val_x] + 1  
}
rownames(mat.temp) <- 1:20; colnames(mat.temp) <- 1:20

pheatmap(mat.temp, cluster_rows=FALSE, cluster_cols=FALSE,
         display_numbers=TRUE, number_format = "%.0f",
             main="Frequency of States at Positions 1 and 2")
```

```{r, fig.width=7, fig.height=7}
idx_x <- 3 # corresponding to the position going into the x axis
idx_y <- 4 # corresponding to the position going into the y axis

mat.temp <- matrix(0, nrow=20, ncol=20)
for (i in 1:400){
  val_x <- data.raw[i,idx_x] 
  val_y <- data.raw[i,idx_y] 
  mat.temp[val_y,val_x] <- mat.temp[val_y,val_x] + 1  
}
rownames(mat.temp) <- 1:20; colnames(mat.temp) <- 1:20

pheatmap(mat.temp, cluster_rows=FALSE, cluster_cols=FALSE,
         display_numbers=TRUE, number_format = "%.0f",
             main="Frequency of States at Positions 3 and 4")
```


## Functions

#### `get_CM`: function to obtain compressed precision matrix (CPM) and compressed covariance matrix (CCM)

```{r}
# input: L, F, E     where matrix X=LF+E
#        data column index as [position, state] matrix
# output: compressed precision matrix (CPM) and compressed covariance matrix (CCM)
get_CM <- function(l,f,e, col.idx){
  
    
    #########################
    # Compressed Precision Matrix
  
    # exclude first factor (which captures mean level) and loadings for numerical stability
    #apply(l,2,sd) # check standard deviation of the loadings by factors
    l2 <- l[,-1]
    f2 <- f[-1,]
    
    Psi <- cov(e) # error covariance
    Psi.inv <- diag(diag(Psi)^{-1})
    
    Lambda.L <- cov(l2)
    Lambda.L.inv <- diag(diag(Lambda.L)^{-1})
    Omega <- Psi.inv - Psi.inv %*% t(f2) %*% solve(Lambda.L.inv+f2%*%Psi.inv%*%t(f2)) %*% f2 %*% Psi.inv
  
    # measure position i -- position j interaction as sqrt(sum of squares of Omega_{k, l})
    #         where position(k)=i and position(l)=j
    data.frame(value = c(Omega),
               position1 = rep(col.idx[,1], times=nrow(col.idx)),
               position2 = rep(col.idx[,1], each=nrow(col.idx))
               ) %>%
      group_by(position1, position2) %>%
      summarise(value = sqrt(sum(value^2))) -> sumsq
    
    # compressed precision matrix
    matrix(sumsq$value, byrow=FALSE, 
           ncol=length(unique(col.idx[,1])) # = number of positions
           ) -> CPM
    
    
    #########################
    # Compressed Covariance Matrix
    Sigma <- t(f) %*% cov(l) %*% f + cov(e)
    
    # measure position i -- position j interaction as sqrt(sum of squares of Omega_{k, l})
    #         where position(k)=i and position(l)=j
    data.frame(value = c(Sigma),
               position1 = rep(col.idx[,1], times=nrow(col.idx)),
               position2 = rep(col.idx[,1], each=nrow(col.idx))
               ) %>%
      group_by(position1, position2) %>%
      summarise(value = sqrt(sum(value^2))) -> sumsq
    
    # compressed precision matrix
    matrix(sumsq$value, byrow=FALSE, 
           ncol=length(unique(col.idx[,1])) # = number of positions
           ) -> CCM
    
    
    out.list <- list(CPM=CPM, CCM=CCM)
    
    return(out.list)
}
```


#### `plot_CM`: function to plot compressed precision matrix (CPM) or compressed covariance matrix (CCM)

```{r}
# input: compressed precision matrix (output from `get_CPM`), 
#        cutoff: to make the figure to easy to check, do not show cells with values < cutoff
# output: heatmap of compressed precision/covariance matrix (dim = #positions X #positions)

plot_CM <- function(CM, cutoff=0, type=1){
  
    # plot only off-diagonal elements (otherwise, diagonal elements dominate visually)
    diag(CM) <- NA
    CPM[CM<cutoff] <- NA
    
    if(type==1){pheatmap(CM, cluster_rows=FALSE, cluster_cols=FALSE,
             main="Compressed Precision Matrix")}
    if(type==2){pheatmap(sqrt(CM), cluster_rows=FALSE, cluster_cols=FALSE,
              main="Compressed Precision Matrix: sqrt transformed")}
    if(type==3){pheatmap(log(CM), cluster_rows=FALSE, cluster_cols=FALSE,
                    main="Compressed Precisoin Matrix: log transformed")}
}

```



## Estimations

```{r}
library(flashier)
# maximum number of factors to be added
fmax <- 300

tic()
fit <- flashier::flash(data, backfit=TRUE, verbose.lvl = 0, greedy.Kmax = fmax, 
                       prior.family = prior.point.normal(),
                       var.type=2) # column-specific error variance
toc() # takes around 90 seconds

# calculate l, f, e
l <- fit$loadings.pm[[1]]
f <- t(fit$loadings.pm[[2]])
e <- data - (l %*% f) 

# obtain compressed precision matrix
list.CM <- get_CM(l, f, e, col.idx)
CPM <- list.CM$CPM
CCM <- list.CM$CCM
```

101 factors were fitted initially, 9 out of which were removed in the backfitting procedure. For the final fit, we have 92 factors.





## Results

### The Truth

This is the `truth` of the underlying interactions. This corresponds to the leftmost plot of the Figure 5 in Ingraham and Marks (2017).

```{r truth, fig.width=4, fig.height=4}
mat.truth <- truth.location
diag(mat.truth) <- NA
pheatmap(mat.truth, cluster_rows=FALSE, cluster_cols=FALSE,
         main="True Interactions")
```

### Compressed Precision Matrix conditional on Truth values

```{r CPM_conditional, fig.width=8, fig.height=4}
plot_list <- list()
breaksList = seq(-3.5, 7.5, by = 0.5)
colpal <- colorRampPalette(rev(brewer.pal(n =10, name = "RdBu")))(length(breaksList))

mat.temp <- CPM
diag(mat.temp) <- NA
mat.temp[mat.truth!=1] <- NA
pheatmap(log(mat.temp), cluster_rows=FALSE, cluster_cols=FALSE,border_color='white',
         color=colpal,breaks = breaksList, silent=TRUE,
         main="CPM: Truth==1, log transformed") -> temp.fig
plot_list[[1]] <- temp.fig[[4]]

mat.temp <- CPM
diag(mat.temp) <- NA
mat.temp[mat.truth!=0] <- NA
pheatmap(log(mat.temp), cluster_rows=FALSE, cluster_cols=FALSE,border_color='white',
         color=colpal,breaks = breaksList, silent=TRUE,
         main="CPM: Truth==0, log transformed") -> temp.fig
plot_list[[2]] <- temp.fig[[4]]

grid.arrange(arrangeGrob(grobs= plot_list,ncol=2))
```


### Compressed Covariance Matrix conditional on Truth values

```{r CCM_conditional, fig.width=8, fig.height=4}
plot_list <- list()
breaksList = seq(-5.5, -0.5, by = 0.5)
colpal <- colorRampPalette(rev(brewer.pal(n =10, name = "RdBu")))(length(breaksList))
plot_list <- list()

mat.temp <- CCM
diag(mat.temp) <- NA
mat.temp[mat.truth!=1] <- NA
pheatmap(log(mat.temp), cluster_rows=FALSE, cluster_cols=FALSE,border_color='white', 
         color=colpal,breaks = breaksList, silent=TRUE,
         main="CCM: Truth==1, log transformed") -> temp.fig
plot_list[[1]] <- temp.fig[[4]]

mat.temp <- CCM
diag(mat.temp) <- NA
mat.temp[mat.truth!=0] <- NA
pheatmap(log(mat.temp), cluster_rows=FALSE, cluster_cols=FALSE,border_color='white', 
         color=colpal,breaks = breaksList, silent=TRUE,
         main="CCM: Truth==0, log transformed") -> temp.fig
plot_list[[2]] <- temp.fig[[4]]

grid.arrange(arrangeGrob(grobs= plot_list,ncol=2))
```



### True Positive Ratio

This is the true positive ratio (for the `truth` of the underlying interactions). This corresponds to the rightmost plot of the Figure 5 in Ingraham and Marks (2017). Along with the flashier-based prediction shown in red, the line corresponding to perfect prediction (where first 100 predictions are the true interactions) and the line corresponding to random prediction (where the true positive ratio is fixed at 100/1225, which is the ratio of true positives) are shown.

```{r}
data.frame(cpm = CPM[upper.tri(CPM, diag=FALSE)],
           ccm = CCM[upper.tri(CCM, diag=FALSE)],
           truth = truth.location[upper.tri(truth.location, diag=FALSE)]
           ) -> df.temp
df.temp %>% arrange(-cpm) %>%
  mutate(cumsum_truth = cumsum(truth),
         nrow=row_number(),
         true_positive = cumsum_truth/nrow)-> df.temp1
df.temp %>% arrange(-ccm) %>%
  mutate(cumsum_truth = cumsum(truth),
         nrow=row_number(),
         true_positive = cumsum_truth/nrow)-> df.temp2

ggplot()+
  geom_line(aes(x=1:nrow(df.temp1), y=df.temp1$true_positive, col='CPM'))+
  geom_line(aes(x=1:nrow(df.temp2), y=df.temp2$true_positive, col='CCM'))+
  geom_line(aes(x=1:nrow(df.temp1), y=c(1:100, rep(100,times=nrow(df.temp)-100))/(1:nrow(df.temp)), 
                col='perfect prediction'))+
  geom_line(aes(x=1:nrow(df.temp1), y=100/1225, col='random prediction(=100/1225)'))+
  ylim(0,1)+ylab("")+
  xlab("Top N interactions")+ ggtitle("True Positive Ratio")
```

This is the true positive ratio plot for CCM and first 120 interactions.

```{r, fig.width=4, fig.height=4}
ggplot()+
  geom_line(aes(x=1:120, y=df.temp2$true_positive[1:120], col='CCM'))+
  ylim(0.5,1)+ylab("")+scale_x_continuous(limits=c(0,120),breaks=seq(0,120,by=20))+
  xlab("Top N interactions")+ ggtitle("True Positive Ratio")
```



